The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-09 paper/ms.tex
small changes based on other Chervin comments

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not
entirely due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar in mass to
Sagittarius, which has punched through the galactic disk outside the solar
circle, leaving behind some of its stars and inducing warping in the disk.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider
results from a live N-body simulation of a Sagittarius-like encounter. This
simulation offers us the ability to see how the midplane varies in a more
controlled environment. The details of the simulation are given in
\citet{2018MNRAS.481..286L}, but we briefly summarize the most relevant
details here. For the Milky Way, the dark halo is modeled as a Hernquist
sphere of mass $10^{12}\,\Msun$ and scale length of $52\,\kpc$, the disk
modeled as an exponential disk with a scale radius of $3.5\,\kpc$, scale
height $0.53\,\kpc$, and mass $6\times10^{10}\,\Msun$, and a Hernquist bulge
of mass $10^{10}\,\Msun$ and scale length $0.7\,\kpc$
\citep{1990ApJ...356..359H}. We use the L2 model, for which Sagittarius is
modeled as a dark matter Hernquist sphere of mass $8\times10^{10}\,\Msun$ and
scale length $8\,\kpc$, along with a stellar component modeled as a Hernquist
sphere of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All
components are realized with live distributions of N-body particles; the Milky
Way and Sagittarius are each initialized to be in equilibrium in isolation
\Gus{check!}.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur
which bring the Sagittarius-like object closer to the Solar circle, we can see
that stronger midplane variations result. In fact, one can see that an $m=2$
mode is excited in the lower left panel at $t=6.0\,\Gyr$ --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The fact that the $t=6.9\,\Gyr$
panel looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that the midplane variations are
produced in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the MW with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim185, 162, 84\,\pc$ (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim30$ and $\sim100$ pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the MW suggests that these values are consistent with expectations for midplane variations in the MW itself, which have not yet been measured as a function of azimuth.

@@

\item Some inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely important for some
applications. In particular, it is imperative to test the assumption that the
dynamical Galactic center is colocated with \sgra{}. We discuss how to
do this in \S \ref{ssec:coord_off}.

\item This work emphasizes the importance of combining chemistry and dynamics.
Chemical tagging \citep{2002ARA&A..40..487F} and dynamical tagging must
complement and confirm each other. The importance of combining chemistry and
dynamics for Galactic archaeology was recently pointed out by
\citet{2019arXiv190210719K}.

>>>>

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not necessarily due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar in mass to
Sagittarius, which has punched through the galactic disk outside the solar
circle, leaving behind some of its stars and inducing warping in the disk.

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider
results from a live N-body simulation of a Sagittarius-like encounter. This
simulation offers us the ability to see how the midplane varies in a more
controlled environment. The details of the simulation are given in
\citet{2018MNRAS.481..286L}, but we briefly summarize the most relevant
details here. For the Milky Way, the dark halo is modeled as a Hernquist
sphere of mass $10^{12}\,\Msun$ and scale length of $52\,\kpc$, the disk
modeled as an exponential disk with a scale radius of $3.5\,\kpc$, scale
height $0.53\,\kpc$, and mass $6\times10^{10}\,\Msun$, and a Hernquist bulge
of mass $10^{10}\,\Msun$ and scale length $0.7\,\kpc$
\citep{1990ApJ...356..359H}. We use the L2 model, for which Sagittarius is
modeled as a dark matter Hernquist sphere of mass $8\times10^{10}\,\Msun$ and
scale length $8\,\kpc$, along with a stellar component modeled as a Hernquist
sphere of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All
components are realized with live distributions of N-body particles; the Milky
Way and Sagittarius are each initialized to be in equilibrium in isolation.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur
which bring the Sagittarius-like object closer to the Solar circle, we can see
that stronger midplane variations result. In fact, one can see that an $m=2$
mode is excited in the lower left panel at $t=6.0\,\Gyr$ --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The fact that the $t=6.9\,\Gyr$
panel looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) suggests that midplane variations are
produced at least in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the MW with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim$185, 162, 84 pc (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim$30 and $\sim$100 pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the MW suggests that these values are consistent with expectations for midplane variations in the MW itself, which have not yet been measured as a function of azimuth.

@@

\item Inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely important for some
applications. In particular, it is imperative to test the assumption that the
dynamical Galactic center is colocated with \sgra{}. We discuss how to
do this in \S \ref{ssec:coord_off}.

\item This work underlines the importance of combining chemistry and dynamics.
Since chemical tagging \citep{2002ARA&A..40..487F} is not subject to the same systematic errors discussed in this paper, it should be used to confirm dynamical associations and to offset the effect of these systematic errors on the action-space distribution function. %The importance of combining chemistry and dynamics for Galactic archaeology was recently pointed out by \citet{2019arXiv190210719K}.
